taxize - taxonomic search and retrieval in R

### 2013-07-19 inst/doc/taxize_withcode_plos.tex

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
All species are hierarchically related to one another, and we use taxonomic names to label the nodes in this hierarchy. Taxonomic data is becoming increasingly available on the web, but scientists need a way to access taxonomic data on the web in a programmatic fashion that's easy and reproducible. We have developed taxize, an open-source software package (freely available from \url{http://cran.r-project.org/web/packages/taxize/index.html}) for the R language. taxize provides simple, programmatic access to taxonomic data for 13 data sources around the web. We discuss the need for a taxonomic toolbelt in R, and outline a suite of use cases for which taxize is ideally suited (including a full workflow as an appendix). The taxize package facilitates open and reproducible science by allowing taxonomic data collection to be done in the open-source R platform.

@@

The use of taxonomic names is, unfortunately, not straightforward. Taxonomic names often change due to name changes at the generic or specific levels, lumping or splitting lower taxa (genera, species) among higher taxa (families), and name spelling changes. For example, a study found that a compilation of 308,000 plant observations from 51 digitized herbarium records had 22,100 unique taxon names, of which only ~13,000 were accepted names \cite{weiser2007,boyle2013}. The scale of this study may be larger than most, but the problem persists for any size study. In addition, there is no one authoritative taxonomic names source for all taxa - although, there are taxon specific sources that many scientists use that study that taxon. Different sources (e.g., uBio, Tropicos, ITIS) may have different accepted names for the same taxon. For example, while the Integrated Taxonomic Information Service (ITIS) has \emph{Helianthus x glaucus} as an accepted name, The Plant List \url{http://www.theplantlist.org} has that name as unresolved, but has \emph{Helianthus glaucus} as an accepted name, while ITIS doesn't have the name.

@@

Taxonomic data is getting easier to obtain through web interfaces (e.g., \cite{eol}). However, there are a number of good reasons to obtain taxonomic information programatically rather than through a web interface. First, if you have more than a few names to lookup on a website, it can take quite a long time to enter each name, get data, and repeat for each species. Second, programatically getting taxonomic names solves the first problem by looping over a list of names. In addition, doing taxonomic searching, etc. is reproducible. With increasing reports of irreproducibility in science \cite{stodden2010,zimmer2012}, it is extremely important to make science workflows repeatable. Science workflows can now easily incorporate text, code, and images in a single executable document \cite{yihui2013}.

The R language is a widely used language by biologists, and now has over 4,000 packages on the Comprehensive R Archive Network (CRAN) and more than 2,500 packages on other repositories to extend R. R is great for manipulating, visualizing and fitting statistical models to data. However, the key missing piece in R is the ability to get data from the internet within R. Getting data from the web will be increasingly common as more and more data gets moved to the cloud. Increasingly, data is available from the web via application programming interfaces (API). These are bits of code that allow computers to talk to one another using code that is not human readable, but is machine readable. Web APIs often define a number of methods that allow users to search for a species name, or retrieve the synonyms for a species name, for example. A further advantage of APIs is that they are language agnostic, meaning that data can be consumed in almost any computing context, allowing users to interact with the web API without having to know the details of the code. Moreover data can be accessed from every computer, whereas for example data stored in an Excel file can only be opened in a few programs.

The goal of taxize, an R package in development, is to make all use cases having to do with retrieving and resolving taxonomic names easy and reproducible. In taxize, we have written a suite of R functions that interact with many taxonomic data sources via their web APIs (Table~\ref{tab:a}). The interface to each function is usually a simple list of species names, just as a user would do with a web API. Therefore, we hope moving from a web to R interface for taxonomic names will be relatively seamless (if one is already nominally familiar with R).

@@

New data sources can be added; we may add the following sources: Wikispecies and The Tree of Life. A connection to \url{freshwaterecology.info} \cite{freshwaterecology} (a database with autecological characteristics, ecological preferences and biological traits as well as distribution patterns of more than 12.000 European freshwater organisms belonging to fish, macro-invertebrates, macrophytes, diatoms and phytoplankton) will be finished when their new API will be released.
In addition, the authors may be contacted for further suggestions of data sources to be added.

@@

This is a common task in biology. We often have a list of species names and we want to know if a) we have the most up to date names, b) our names are spelled correctly, and c) if we have common names, we likely need the scientific names. One way to resolve names is via the Global Names Resolver (GNR) service provided by the Encyclopedia of Life \cite{eolgnr}. Here, we are searching for two misspelled names:

@@

taxize takes the approach that the user should be able to make decisions about what resource to trust, rather than taxize making the decision. Both the EOL GNR and the TNRS services provide data from a variety of data sources. The user may trust a specific data source, thus may want to use the names from that data source. In the future, we may provide the ability for taxize to suggest the best match from a variety of sources, but since R is relatively inefficient in memory management, etc., we would rather offload this sort of computationally intensive task.

@@

A number of data sources in taxize provide the capability to retrieve higher taxonomic names, but we will highlight two of the more useful ones: Integrated Taxonomic Information System (ITIS) \cite{itis} and National Center for Biotechnology Information (NCBI) \cite{federhen}. First, we'll search for two species, \emph{Abies procera} and \emph{Pinus contorta}.

@@

It turns out that the panther has a status of endangered (EN) and the lynx has a status of least concern (LN).

>>>>

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
All species are hierarchically related to one another, and we use taxonomic names to label the nodes in this hierarchy. Taxonomic data is becoming increasingly available on the web, but scientists need a way to access it in a programmatic fashion that's easy and reproducible. We have developed taxize, an open-source software package (freely available from \url{http://cran.r-project.org/web/packages/taxize/index.html}) for the R language. taxize provides simple, programmatic access to taxonomic data for 13 data sources around the web. We discuss the need for a taxonomic toolbelt in R, and outline a suite of use cases for which taxize is ideally suited (including a full workflow as an appendix). The taxize package facilitates open and reproducible science by allowing taxonomic data collection to be done in the open-source R platform.

@@

The use of taxonomic names is, unfortunately, not straightforward. Taxonomic names often vary due to name changes at the generic or specific levels, lumping or splitting lower taxa (genera, species) among higher taxa (families), and name spelling changes. For example, a study found that a compilation of 308,000 plant observations from 51 digitized herbarium records had 22,100 unique taxon names, of which only ~13,000 were accepted names \cite{weiser2007,boyle2013}. In addition, there is no one authoritative taxonomic names source for all taxa - although, there are taxon specific sources that are used by many scientists. Different sources (e.g., uBio, Tropicos, ITIS) may use different accepted names for the same taxon. For example, while the Integrated Taxonomic Information Service (ITIS) has \emph{Helianthus x glaucus} as an accepted name, The Plant List \cite{theplantlist} has that name as unresolved. But \emph{Helianthus glaucus} is an accepted name in The Plant List, while ITIS does not list this name.

@@

Taxonomic data is getting easier to obtain through the web (e.g., \cite{eol}). However, there are a number of good reasons to obtain taxonomic information programatically rather than through a web interface. First, if you have more than a few names to lookup on a website, it can take quite a long time to enter each name, get data, and repeat for each species. Programatically getting taxonomic names solves the problem by looping over a list of names. In addition, doing taxonomic searching, etc. becomes reproducible. With increasing reports of irreproducibility in science \cite{stodden2010,zimmer2012}, it is extremely important to make science workflows repeatable. Science workflows can now easily incorporate text, code, and images in a single executable document \cite{yihui2013}.

The R language is a widely used language by biologists, and now has over 4,000 packages on the Comprehensive R Archive Network (CRAN) and more than 2,500 packages on other repositories to extend R. R is great for manipulating, visualizing and fitting statistical models to data. \cite{gentleman_bioconductor:_2004} gives a detailed discussion of advantages of R in computational biology. Getting data from the web will be increasingly common as more and more data gets moved to the cloud. Therefore there is a need to get data from the web directly into R. Increasingly, data is available from the web via application programming interfaces (API). These allow computers to talk to one another using code that is not human readable, but is machine readable. Web APIs often define a number of methods that allow users to search for a species name, or retrieve the synonyms for a species name, for example. A further advantage of APIs is that they are language agnostic, meaning that data can be consumed in almost any computing context, allowing users to interact with the web API without having to know the details of the code. Moreover data can be accessed from every computer, whereas for example an Excel file can only be opened in a few programs.

The goal of taxize, an R package in development, is to make many use cases having to do with retrieving and resolving taxonomic names easy and reproducible. In taxize, we have written a suite of R functions that interact with many taxonomic data sources via their web APIs (Table~\ref{tab:a}). The interface to each function is usually a simple list of species names, just as a user would do with a web API. Therefore, we hope moving from a web to R interface for taxonomic names will be relatively seamless (if one is already nominally familiar with R).

@@

New data sources can be added; we plan to add the following sources: Wikispecies and The Tree of Life. A connection to \url{freshwaterecology.info} \cite{freshwaterecology} (a database with autecological characteristics, ecological preferences and biological traits as well as distribution patterns of more than 12.000 European freshwater organisms belonging to fish, macro-invertebrates, macrophytes, diatoms and phytoplankton) will be finished when their new API will be released.
In addition, the authors may be contacted for further suggestions of data sources to be added.

@@

This is a common task in biology. We often have a list of species names and we want to know a) if we have the most up to date names, b) if our names are spelled correctly, and c) the scientific name for a common name. One way to resolve names is via the Global Names Resolver (GNR) service provided by the Encyclopedia of Life \cite{eolgnr}. Here, we are searching for two misspelled names:

@@

taxize takes the approach that the user should be able to make decisions about what resource to trust, rather than making the decision. Both the EOL GNR and the TNRS services provide data from a variety of data sources. The user may trust a specific data source, thus may want to use the names from that data source. In the future, we may provide the ability for taxize to suggest the best match from a variety of sources.

@@

A number of data sources in taxize provide the capability to retrieve higher taxonomic names, but we will highlight two of the more useful ones: Integrated Taxonomic Information System (ITIS) \cite{itis} and National Center for Biotechnology Information (NCBI) \cite{federhen}. First, we'll search for two species, \emph{Abies procera} and \emph{Pinus contorta} within ITIS.

@@

It turns out that the panther has a status of endangered (EN) and the lynx has a status of least concern (LC).
